---
title: "Fertility Exercise 3: Simulating a moving target"
output: html_notebook
---

The goal of this exercise is see how period fertility reacts to a small change in desired completed family size. 

The exercises have 3 parts:

1. Simulate  single cohort with a single target. This reveals an exponential solution to duration-specific fertility.

2. Simulate multiple cohorts on the Lexis surface -- still with a constant target. (This mostly just illustrates how to use loops to do this.)

3. Simulate multiple cohorts with a period change in targets. 

## 1. Simulate fixed target for a cohort

The code below will recursively simulate the differential equation
$$
        f_x = \alpha * (D - F_x)
$$        

```{r}
## We'll assume 10 age groups
n_ages = 10
Fx = rep(0, n_ages) ## initialize cum fert at 0
fx = rep(NA, n_ages) ## and duration-specific fert as NA
alpha = 1/3 ## fulfillment rate
D = 1 ## target (1 child)

for (i in 1:n_ages) ## loop thru all ages
{
    if (i == 1) ## special case where cumulative fertility is 0
        {
            fx[i] = alpha * (D - 0)
            Fx[i] <- 0    + fx[i] ## assign first cumu fertility thru first age group
        }
    if (i > 1) ## general case
    {
        fx[i] = alpha * (D - Fx[i-1]) ## estimate fertility at duration i
        Fx[i] <- Fx[i-1] + fx[i] ## update cumulative fertility
    }
}
```

To visualize

```{r}

par(mfrow = c(2,1))
plot(fx, type = 'o',
     main = "Duration-specific fertility",
     xlab = "Duration",
     axes = F)
axis(1); axis(2, at = round(fx,2))
plot(Fx, type = 'o',
     main = "Cumulative fertility",
     xlab = "Duration")
```

Discussion questions:
1a. Does cumulative cohort fertility eventally reach the target?
1b. Does duration-specific fertility seem realistic?


## 2. Put into cohorts on a lexis surface and calculate period TFR
## still with a constant desired fertility target of 1.0

We want to simulate
$$
        f_{x,t} = \alpha * (D_t - F^{coh}_{x,t})
$$   
```{r}
n_ages = 10
n_per = 20
Dt <- rep(1, n_per) ## a vector of life-time desires, 1 for all periods

## now do a double loop over lexis surface
fxt <- matrix(NA, n_ages, n_per)
Fxt <- matrix(NA, n_ages, n_per)
## initialize cumulative fertility 
Fxt[,1] <- Fx # steady state for starting period
fxt[,1] <- fx # steady state for starting period
for (i in 1:n_ages)
{
    for (j in 2:n_per) # start in 2nd period (already assigned 1st)
    {
        if (i == 1) ## youngest age has 0 cumulative fertility
        {
            fxt[i,j] = alpha * (Dt[j] - 0)
            Fxt[i,j] <- 0    + fxt[i,j]
        }
        if (i > 1) ## general case, we just update as before, but for each cohort
        {
            fxt[i,j] = alpha * (Dt[j] - Fxt[i-1,j-1])
            Fxt[i,j] <- Fxt[i-1,j-1] + fxt[i,j]
        }
    }
}
```

Visualize results
```{r}
tfr = apply(fxt, 2, sum)
plot(tfr)
```

Discussion questions: 
2a. Why did we have to initialize 1st columns of fxt and Fxt
2b. What happens to period TFR? Nothing, right?

## 3. Now simulate a 1-time, permanent increase in target from 1.0 to 1.2

Same code as before with 1 line changed
```{r}
n_ages = 10
n_per = 20
## Dt <- rep(1, n_per)
Dt <- c(rep(1, n_per/2),  rep(1.2, n_per/2)) ## New line!!!
print(Dt)        

## now do a double loop over lexis surface
fxt <- matrix(NA, n_ages, n_per)
Fxt <- matrix(NA, n_ages, n_per)
## initialize cumulative fertility 
Fxt[,1] <- Fx # steady state for starting period
fxt[,1] <- fx # steady state for starting period
for (i in 1:n_ages)
{
    for (j in 2:n_per) # start in 2nd period (already assigned 1st)
    {
        if (i == 1) ## youngest age has 0 cumulative fertility
        {
            fxt[i,j] = alpha * (Dt[j] - 0)
            Fxt[i,j] <- 0    + fxt[i,j]
        }
        if (i > 1)
        {
            fxt[i,j] = alpha * (Dt[j] - Fxt[i-1,j-1])
            Fxt[i,j] <- Fxt[i-1,j-1] + fxt[i,j]
        }
    }
}

tfr = apply(fxt, 2, sum)
plot(tfr)
```

3a. When is period tfr highest? Why do you think this is?
3b. What does period tfr converge to at the end?
3c. Does this help explain covid slide about "boom", "bust" and "thud"?                                                                          

## 4. Optional extra credit. Implement 1-time decrease in fulfillment $\alpha(t)$.




