---
title: "Fertility Exercise 2: Are recent fertility decline just tempo effects: applying Bongaarts-Feeney in South Korean and Finland"
author: "Joshua R. Goldstein"
date: "Berkeley workshop, 2023"
output: html_notebook
---

In these exercises, we will look at recent fertility change in several countries and assess the extent to which the changing pace of postponement may be influencing period total fertility

The exercises is mostly just running the existing code and interpreting the output.

We will use data from www.humanfertility.org, which we've already put on our github page for easy downloading.

## Preliminaries

We start by downloading some utility functions and loading our fertility data. 

```{r}
source("fertility_exercise_functions.R")

asfr_file = "asfrRR.txt"
```

## Creating a matrix of age-specific fertility rates

```{r}
A <- get_hfd_fertility_matrix(file = asfr_file, code = "KOR")
A <- A[,paste(2000:2020)] ## just keep years 2000 to 2020


per = as.numeric(colnames(A))
age = as.numeric(rownames(A))

## get period TFR
per_tfr = apply(A, 2, sum)
```

## Visualize TFR, age-specific rates, and adjusted-TFR

The code below will produce a set of panels in a single plot.

```{r}
par(mfrow = c(3,2))

## plot TFR
plot(per, per_tfr, type = 'o')  ## you might want to adjust ylim argument in plot()
title("Total Fertility Rate")
grid()

## plot ASFRs
A5 <- A[age %in% seq(15, 45, 5),]
matplot(per, t(A5), type = 'l',
        ylab = "asfr",
        xlab = "period")
title("Age-specific fertility trends")
text(rep(per[1], nrow(A5)),
     A5[,1], rownames(A5), col = 1:6)


## now calculate bongaarts-feeney (without parity)

mu_per = apply(A, 2, get.mean)
mu_per_prime = center.diff(mu_per, end.fill = T)

## plot Mean Age
plot(per, mu_per)
title("Mean age of childbearing")
grid()

## plot rate of change in Mean Age
plot(per, mu_per_prime, ylim = c(0, 0.3));
title("Rate of postponement")
grid()

## calculate bongaarts-feeney adjusted TFR
adjusted_per_tfr <- per_tfr / (1 - mu_per_prime)

## plot adjusted TFR
plot(per, per_tfr, ylim = c(0.7, 2.2), pch = 19)
grid()
points(per, adjusted_per_tfr, type = "p", col = "red")

## Finally, et's get the cohort turning 20 in 2000 and see how many kids they had by 40

per = as.numeric(colnames(A))
age = as.numeric(rownames(A))
Per = per[col(A)]
Age = age[row(A)]
Coh = Per-Age

my_cohort_asfr = A[Coh == 1980]
my_ctfr = sum(my_cohort_asfr) ## [1] 1.43224 

## add horizontal line to TFR plot showing this cohort
abline(h = my_ctfr, col = "orange", lty = 2)
title("Period TFR, observed and adjusted")
## add legend
legend("bottomleft",
       legend = c("TFR", "TFR-adj", "1980 cohort"),
       pch = c(19, 1, -1),
       col = c("black", "red", "orange"),
       lty = c(-1,-1, 1))
```

Questions;
1. How low does TFR get? What long-term growth rate would this imply if it stayed constant for many decades?
2. Are age-specific changes linear enough to use Ryder translation equation accurately?
3. About how much bigger in percentage terms is the tempo-adjusted period TFR than the unadjusted one? (Hint: look at the mu_per_prime graph).
4. Does faster postponement seem to explain the recent period TFR declines?

Extra-credit: repeat exercise for Finland, country code "FIN".

Extra-extra-credit: look up published Bongaarts-Feeneey tempo-adjusted TFRs on www.humanfertility.org and see if these -- which incorporate separate estimates by parity -- show same pattern as what we found here.





