---
  title: "Fertility Exercise 1: Translating from period to cohort"
author: "Joshua R. Goldstein"
date: "Berkeley Workshop, 2023"
output: html_notebook
---

  In these exercises, we will apply the version of Ryder's translation that converts period total fertility to cohort total fertility.

  
$$
CTFR\left(t - \mu_p(t)\right) = PTFR(t) \times \left[1 + \mu'_p(t) \right]
$$
  * Notice that all parameters involving the mean are in terms of easily measurable period quantities.
  * Remember that this equation is only exactly true when age-specific fertility is changing perfectly linearly; otherwise, it is only an approximation.
In the exercises we will look at:
  1. Our toy example from the pre-workshop exercises
  2. Real data from the United States baby-boom and bust

## Pre-workshop exercise (cont.)
Here is what we already did:
```{r}
## input matrix of fertility rates
## by age (rows) and period (columns)
A <- Fxt <- matrix(
  data = c(1.0,  0.9,  0.8,  0.7,  0.6,  0.5,
           1.0,  1.0,  1.0,  1.0,  1.0,  1.0,
           0.5,  0.6,  0.7,  0.8,  0.9,  1.0),
  nrow = 3,
  ncol = 6,
  byrow = TRUE)
## define indices for Age, Period, and Cohort
Age = row(A)
Per = col(A)
Coh = Per - Age ## Since, cohort  + age = period
cohort_tfr = tapply(X = A, INDEX = Coh, FUN = sum)
period_tfr = tapply(X = A, INDEX = Per, FUN = sum)
print(cohort_tfr)
print(period_tfr)
## save period and cohort values
period = as.numeric(names(period_tfr))
cohort = as.numeric(names(cohort_tfr))
```

Now we'll use Ryder's translation to estimate the cohort TFR from the period quantities.

```{r}
table_mean <- function(Fx, x.mid)
{
  mu = sum(x.mid * Fx)/sum(Fx)
  return(mu)
}
## get period mean ages
my_x.mid = c(0.5, 1.5, 2.5)
mu_p = tapply(X = A,
              INDEX = Per,
              FUN = table_mean,
              x.mid = my_x.mid)
print(mu_p)
mu_p_prime = c(NA, diff(mu_p)) ## diffs for last interval
cohort_tfr_hat = period_tfr * (1 + mu_p_prime) ## key formula
cohort_hat = period - mu_p ## corresponding birth cohort for each cohort_tfr
## plot the results
par(mfrow = c(1,1))
plot(period, period_tfr,
     col = "blue", xlim = range(c(cohort, period)),
     ylim = c(0,3), pch = 19,
     xlab = "period for ptfr or birth cohort for ctfr")
title("Period and Cohort Total Fertility for Pre-workshop Example")
lines(cohort, cohort_tfr,
      type = "p", col = "red", pch = 19)
points(cohort_hat, cohort_tfr_hat, col = "red")
arrows(x0 = period, x1 = cohort_hat, y0 = period_tfr, y1 = cohort_tfr_hat, col = "grey", length = 0.1)
legend("topleft", legend = c("period", "cohort", "cohort est."),
       col = c("blue", "red", "red"),
       pch = c(19, 19, 1), bty = "n")
```

Questions:
  1. Which cohort TFRs should we take seriously? Which should we ignore? Why?
  #A: In this example: 
  2. How well does Ryder estimate work in matching cohort TFRS? (a) ok, (b) perfectly, (c) not very well.

## Part 2: Real US data
Here we'll do the cohort born in about 1970, having kids (on average) in about year 2000.
```{r}
## We'll read in age-specific fertility rates from Human Fertility Database and put them in a matrix
library(data.table)
dt <- fread("asfrRR.txt")
dt <- dt[Code == "USA"] ## keep just USA
dt[, x := as.numeric(Age)] ## assign age variable as numeric "x"
dt[Age == "12-", x := 12] ## fix youngest and oldest ages
dt[Age == "55+", x := 55]
## put into a matrix using xtabs
A <- dt[, xtabs(ASFR ~ x + Year)]
x = as.numeric(rownames(A))
yr = as.numeric(colnames(A))
Age = x[row(A)]
Per = yr[col(A)]
Coh = Per - Age ## Since, cohort  + age = period
```

Now get TFRs (period and cohort)

```{r}
period_tfr = tapply(X = A, INDEX = Per, FUN = sum)
cohort_tfr = tapply(X = A, INDEX = Coh, FUN = sum)
## drop incomplete cohorts
coh_len = table(Coh)
full_coh = names(coh_len)[coh_len == 44]
cohort_tfr = cohort_tfr[names(cohort_tfr) %in% full_coh]
## save period and cohort values
period = as.numeric(names(period_tfr))
cohort = as.numeric(names(cohort_tfr))
```

To make comparison's easier, we plot cohort TFR at cohort + 28 years, so that each cohort's mean fertility is happening in the period that is being translated. 

```{r}
## get period mean ages
my_x.mid = x + 0.5
mu_p = tapply(X = A,
              INDEX = Per,
              FUN = table_mean,
              x.mid = my_x.mid)
print(mu_p)
mu_p_prime = c(NA, diff(mu_p)) ## diffs for last interval
cohort_tfr_hat = period_tfr * (1 + mu_p_prime)
cohort_hat = period - mu_p
## plot the results
par(mfrow = c(1,2))
plot(period, period_tfr,
     col = "blue", xlim = range(c(cohort+28, period, cohort_hat+28)),
     ylim = c(0,4), pch = 19,
     xlab = "period (or cohort + 28)")
title("Period and Cohort Total Fertility, USA")
lines(cohort + 28, cohort_tfr,
      type = "p", col = "red", pch = 19)
points(cohort_hat + 28, cohort_tfr_hat, col = "red")
legend("topright", legend = c("period", "cohort", "cohort est."),
       col = c("blue", "red", "red"),
       pch = c(19, 19, 1), bty = "n")
##
plot(period, mu_p,
     xlim = range(c(cohort+28, period, cohort_hat+28)))
title("Period mean ages")
grid()
abline(h = 0)
```

Questions:
  1. How does Ryder estimate work? (a) ok, (b) perfectly, (c) not very well.
  2. Can you see any broad pattern in cohort fertility being higher or lower than period fertility? (Note: the answer may be clearer from mean age plot)
  3. Should we believe the translation based on only a period (or 3's) observations? If not, what would be a strategy for getting more robust estimates of periods?
  4. What would you guess that the cohort born in 1995 will have as their cohort fertility: closer to 1.7 or closer to 2.0?